Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types

### 2016-02-04 manuscript.Rnw
new short paragraph in discussion; change of title; edit key paragraph in introduction

\title{Leaf traits and size drive differences in above-ground biomass distribution among woody plant functional types}

@@

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion. 
\begin{itemize}
\item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation. We tested for differences in biomass distribution between major woody plant functional types (PFT), and sought to understand these differences in terms of leaf traits.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($M_F/A_S$) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that $M_F/A_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution, at a given plant size. 
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. \citet{poorter2012} and \citet{enquistniklas2002} attributed higher leaf-stem biomass ratio ($M_F/M_S$) in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms. This is consistent with \citet{poorter2015} who reported lower $M_F/M_S$ in deciduous compared to evergreen species.

It is yet unknown whether plant functional types (PFTs) (angiosperm vs. gymnosperm, deciduous vs. evergreen) with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity. Because longer LL is generally correlated with the leaf mass per area (LMA), PFTs with higher MF/MS (and thus higher LMA) do not necessarily not have larger total leaf area at a given stem mass. Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}. \citet{poorter2015} advanced the hypothesis that $MF/MS$ may be nearly proportional to LMA, arguing that total leaf area is likely less variable between PFTs. Here we set out to directly test this hypothesis with a database that includes observations of leaf area as well as leaf mass measured on the same plants.

@@

We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and the aridity index was calculated as PET/MAP.

@@

From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs. Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with > 1 $m^2$ stem basal area, for which little data was available) (Fig. 4). We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.

@@

We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass. \citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI. Although we found remarkably similar leaf area per unit above-ground woody biomass ($M_F/M_S$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI. To scale our results to stand- and landscape scales requires more information on differences in stand structure between PFTs.

@@

Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height. At a global scale, average plant height is strongly related to climate, possibly via water availability \citet{klein2015,simard2011}. Since biomass distribution itself is steeply related to plant height (Figs. 2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not accounted for. Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.

@@

Leaf area was just about proportional to AS, except for the largest trees. This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise). Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these.

While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.

@@

Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database. We thank Martin De Kauwe and anonymous referees for comments on an earlier version of this paper. Thanks also to Rich FitzJohn for advice about using the remake package. Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. Thanks to a small army of anonymous reviewers for suggestions that greatly improved the paper. Special thanks to Steve Zissou.

>>>>

\title{Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types}

@@

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion. 
\begin{itemize}
\item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation. We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($M_F/A_S$) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that $M_F/A_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution, at a given plant size. 
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. One possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass. If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass. Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}. The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}. Here we set out to directly test the hypothesis that across PFTs, $M_F/M_S$ is related to LMA with a global database that includes observations of leaf area as well as leaf mass measured on the same plants.

@@

We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.

@@

From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs. Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig. 4). We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.

@@

We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass. \citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI. Although we found remarkably similar leaf area per unit above-ground woody biomass ($M_F/M_S$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI. To scale our results to stand- and landscape scales thus requires more information on differences in stand structure between PFTs.

Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}. Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass. It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.

@@

Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height. At a global scale, average plant height is clearly related to climate via water availability \citet{klein2015,simard2011}. Since biomass distribution itself is steeply related to plant height (Figs. 2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not first accounted for. Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.

@@

Global vegetation models assume or predict differences in biomass distribution between PFTs. However, these differences are currently poorly constrained, due to limited available empirical data. Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.

@@

Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database. We thank Martin De Kauwe and anonymous referees for comments on an earlier version of this paper. Thanks also to Rich FitzJohn for advice about using the remake package. Figure 2a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary). Thanks to anonymous reviewers for suggestions that greatly improved the paper. Special thanks to Steve Zissou.
